HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-05 scnn.tex
lots of changes in the result section

% Structure of the section
% 1) Present the goal
% 2) Explain briefly the dataset
% 3) Present the setting
% 4) Describe the result
In this section, we demonstrate the performance of the spherical neural network on a discrimination problem.
Our implementation is available as a small python package \url{https://our.code.link.com}.
Furthermore, all the data to reproduce our experiments are available online.\footnote{\url{https://doi.org/10.5281/zenodo.????} \todo{correct DOI}}.

In our experiment we classify convergence maps, similar to those created with gravitational lensing technique \citep{chang2017curvedsky}.

@@

The simulations are created using the fast lightcone method described in~\citep{sgier2018fastgeneration}.
However, we use only a single simulation box, as opposed to two used in that work, as we use source galaxies at lower redshift of $z=1$, instead of $z=1.5$.
As a simulator we use \pkg{L-PICOLA} \citep{howlett2015lpicola}, a fast and approximate code for N-body simulations.
For each class, we generate 30 simulations.

Out of the 60 simulations, 20 are kept for the test set.
\figref{map_sample} shows the full sky simulations and a zoom region for model 1 (top) and model 2 (bottom).
Initial conditions for these simulations were the same, so the differences in structures can only be attributed to different cosmological parameters used to evolve the particle distribution.

@@

\subsection{Problem formulation}
While the number of maps might sound small for deep learning, the information contained in the maps is very recurrent, allowing us to use parts of the map as sample for our simulation.
Leveraging the properties of the HealPix sampling, we can split each maps
into $12*o^2$ ($o=1,2,4,\dots$) samples that span a smaller part of the sphere. 
As shown in \figref{subpart_sphere}, we used $o=1,2,4$ as the resulting
samples are large enough to suffer from the effects of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

@@

The goal is build a robust classifier, i.e. an automatic way to
discriminate between the two classes in the presence of noise. 
We check the performance of the classifier as a function of the area used, i.e "the order" $o$ and also as a function of the relative level of the noise.

For all the
experiments, we work with centered Gaussian noise. \todo{Tomek: could you put a
justification.} To improve the
stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced during the training process. We trained our
model between \nati{$50$ and $200$} epochs depending on the complexity and the size of
the problem. The testing set was simply used to assess the global performance of
the network and to assess if it was overfitting.
Training took less that a day for each model using a Nvidia 1080 TI.

We compare the performance of the spherical CNN to two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities or b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified use a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better result
while having worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailed in \ref{sec:dataset_augmentation}.

The maps are separated into a training and a testing set. The performance is
assessed by measuring the classification error on a test set. Once the
training maps are sliced into samples, we split them into a training samples
and validation samples. The validation samples is used to tune the
regularization parameter of the SVM using a standard cross-validation
technique.

@@

We present in \figref{results} the results for 5 different noise
levels and $3$ different data sizes.  We observe that the PSD features, i.e.
one of the most used technique for model discrimination in cosmology is unable
to perform the desired task. The main reason is that the two different model
have very similar PSDs (see \figref{psd_sigma3}).   As a result, one
need to use other statistics to solve the problem. In this case, we use
histogram features and observe a significant improvement over the PSD.  The
SCNN present the advantage to learn automatically specific features for the
task and also the usual Deep Learning performance bump.

We also note that the histogram features contains information about the
distribution of pixels, but does not include any spatial information. In
contrast, both the PSD features and the SCNN can exploit the spatial
information. Without this scructure, a non scallable fully connected neural network approach would have to be used.

@@

\subsection{Filter visualization} 
Onecommon visualization for CNN is to display the shape of the learned filters.
Since our construction lead to almost (up to the sampling irregularities)
spherical filters, we plot in \figref{learned_filter} both the section (on the
sphere) and a gnomeview projection on a plane of a random selection of learned
filters. The filters are selected from the last layer of the network
corresponding the experiment with order $o=2$ for a relative noise level of
$2$. Nevertheless, all networks presented similar gnomonic and section pattern, at least from a
human perspective. The gnomonic projection clearly exibits the sampling issue
of HealPix. For a detail, on how the convolution kernels are plot, please refer to
\ref{sec:filter_visualization}. 
\todo{Tomek: can you say something about the filters.}

@@

@@

In this contribution, we presented HealPixNet, a spherical CNN for the HealPix sampling. The gist of our construction reside in using a graph to model the sphere. 
This allows us to leverage the advantages of graph convolution as it is both efficient $\mathcal{O}(n)$ and flexible (allow to work only on a subpart of the sphere) while still well capturing the spherical properties of the domain. 
We also show on a range of classification problems that the neural network systematically significantly outperform classical methods.

We provide HealPixNet as a small easy to use package availlable at \todo{\url{https://github.com/xxx}}. 
As a futur work, we intend to 1) further investigate the link between the graph Fourier basis and the spherical harmonic and 2) compare HealPixNet to other spherical CNN using different sampling schemes.

@@

\section{Spherical harmonics VS graph Fourier eigenvectors}
\ref{fig:graph_harmonics} displays the first 16 eigenvectors of the
graph Fourier basis used for the convolution. While their is a clear
resemblance with the spherical harmonic, the small irregularities in the
sampling (see \figref{healpix_graph_4}) have an effect on the
graph Fourier eigenvectors. First, conterintuitively, some graph eigenvectors will be
localized~\cite{perraudin2018global}, i.e: they will span only a small subpart
of the sphere.  Second, as Nside increase toward infinity and hence the number
of pixel $n\rightarrow \infty$, we are still unsure if the graph eigenvectors
converge toward the spherical harmonics. We believe that it might be the case
for a different graph construction (fully connected) and that a proof could be
built on top of the work of~\cite{belkin2007convergence}. This is obviously
out of the scope of this contribution.

\label{sec:spherical_harmonics}
\begin{figure*}[!htb]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{The 16 first Fourier modes (eigenvectors of the graph Laplacian) of a graph constructed from the Healpix sampling of the sphere. Eigenvectors 1--3 could be associated to spherical harmonics of degree $\ell=1$, eigenvectors 4--8 to degree $\ell=2$, and eigenvectors 9--15 to degree $\ell=3$. Nevertheless, the graph eigenvectors are only approximating the real spherical harmonics.}
\label{fig:graph_harmonics}
\end{figure*}

@@

In \figref{gaussian_filters_visualization}, we present three different
visualization for the a healpix graph convolution kernel. The kernel used is defined in the graph spectral domain as $K_t(x)=e^{-\tau t x}$. This is the domain used for our first visualization. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. The lower eigenvalues $\lambda$ correspond to the lower frequencies, i.e. the spectral mode with less variation. 
As a second visualization, we can observe the kernel directly on the sphere. To do so, we perform the convolution of a
Kroneker on the sphere plot the result with gnomonic projection. Eventually, as we are
working on the sphere with theoretically spherical filters (up the the irregularities of the sampling), we could only plot the section of the convoluted Kroneker. To do so, we convolve a Kroneker on the equator and use the equator for our section as shown in~\figref{index_section}. Note that, because of the small irregularities in the HealPix
sampling, the second and the third methods are subject to small varation depending on the
chosen Kroneker.

@@

When comparing the network, we want to ensure that the SVM classifier has
access to the same amount of data that the network, i.e, potentially an
infinite number of sample. Hence, we fit different linear SVM classifier using
various amount of training set size until we oberve experimentally that
increasing the amount of data does not improve the classifier. The validation
set remain the same and is used for cross-validation to tune the L2
penalization parameter of SVM. Eventually, we evaluate the final classifier on
the testing set. An example of the evolution of the training and valdication
error is show in \figref{hist_error_evolution}.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/hist_error_order2_noise1_5.pdf}
\caption{Evolution curve of the error with respect of the number of training sample for histogram features and a linear classifier (setting: order 2, relative noise level: 1.5). The validation error might be slightly below the training error because we use it for cross-validation.}
\label{fig:hist_error_evolution}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/border_effects.pdf}
\caption{Graph convolution of a Kroneker placed at two different spots on the graph. The graph build from $1/12$ of the sphere and an Nside of 16. When the Kroneker is placed on the left corner, the resulting convolution is not circular anymore due to border effect. The graph setting assume reflective border conditions.}
\label{fig:border_effects}
\end{figure}

>>>>

% Structure of the section
% 1) Present the goal
% 2) Explain briefly the dataset
% 3) Present the setting
% 4) Describe the result
In this section, we demonstrate the performance of the spherical neural network on a discrimination problem. Nevertheless, by simply changing the loss, one could use the network for regression.
Our implementation is available as a small python package \url{https://our.code.link.com}.
Furthermore, all the data to reproduce our experiments are available online.\footnote{\url{https://doi.org/10.5281/zenodo.1303272} \todo{Check DOI}}.

In our experiments we classify convergence maps, similar to those created with gravitational lensing technique \citep{chang2017curvedsky}.

@@

The simulations are created using the fast lightcone method described in~\citep{sgier2018fastgeneration}.
However, we use only a single simulation box, as opposed to two used in that work, as we use source galaxies at lower redshift of $z=1$, instead of $z=1.5$.
As a simulator we use \pkg{L-PICOLA} \citep{howlett2015lpicola}, a fast and approximate code for N-body simulations.
For each class, we generate $30$ simulations (both class have the same initial parameters).

Out of the $60$ simulations, $20$ are kept for the test set.
\figref{map_sample} shows the full sky simulations and a zoom region for model 1 (top) and model 2 (bottom).
Initial conditions for these simulations were the same, so the differences in structures can only be attributed to different cosmological parameters used to evolve the particle distribution.

@@

\subsection{Problem formulation}
While the number of maps might sound small for deep learning, the information contained in the maps is very recurrent, allowing us to use parts of the map as samples for our simulation.
Leveraging the properties of the HealPix sampling, we can split each maps
into $12*o^2$ ($o=1,2,4,\dots$) samples that span a smaller part of the sphere. 
As shown in \figref{subpart_sphere}, we used $o=1,2,4$ as the resulting
samples are large enough to suffer from the effects of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

@@

We build a robust classifier, i.e. an automatic way to
discriminate between the two classes in the presence of noise. 
We check the classification performance as a function of the area used, i.e "the order" $o$ and also as a function of the relative level of the noise added to the map.

For all the
experiments, we work with centered Gaussian noise. \todo{Tomek: could you put a
justification.} To improve the
stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced during the training process. We trained our
model between $50$ and $290$ epochs depending on the complexity and the size of
the problem. $20\%$ of the training set that we call "validation set" was left out of the training process and simply used to assess the global performance and overfitting of
the network.
Training took less than a day for each model using an NVIDIA Tesla K20.

We compare the performance of the spherical CNN against two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities (PSD) and b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified use a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better results
while having worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailed in Appendix~\ref{sec:dataset_augmentation}. We tuned the parameter using cross-validation with the validation set.

@@

We present in \figref{results} the results for 5 different noise
levels and $3$ different sample areas. The spherical CNN has a significant lead for all levels of noise and sample areas. 
In particular, the difference seems to increase as the problem becomes more complex, i.e as the order is increased and there is less information in each sample. 
This is likely due to the fact that the SCNN can learn optimal features for the problem and do not rely on our humanly limited feature design.

We observe that the PSD features, i.e.
one of the most used techniques for model discrimination in cosmology, is unable
to perform the desired task. The main reason is that the two different models
have very similar PSDs (see \figref{psd_sigma3}). As a result, one
needs to use other statistics to solve the problem. In this case, the
histogram features that have been built for the problem allow for a significant improvement over the PSD ones. 
Unfortunately, while the histogram features contain information about the
distribution of pixels, they do not include any spatial information. In
contrast, both the PSD features and the SCNN can exploit the spatial
information. 
Note that without this graph structure, a fully connected neural network approach would have to be used. Problematically, fully connected neural networks do not scale well with the number of pixels.

@@

\subsection{Filter visualization} 
One common visualization for CNN is to observe the shape of the learned filters.
Since our construction leads to almost (up to the sampling irregularities)
spherical filters, we plot in \figref{learned_filter} both the section (on the
sphere) and a gnomonic projection on a plane of a random selection of learned
filters. The filters are selected from the last layer of the network
corresponding the experiment with order $o=2$ for a relative noise level of
$2$. Nevertheless, all networks presented similar gnomonic and section pattern, at least from a
human perspective. The gnomonic projection clearly exibits the sampling issue
of HealPix. For a detail, on how the convolution kernels are plotted, please refer to Appendix~\ref{sec:filter_visualization}. 
\todo{Tomek: can you say something about the filters.}

@@

\nati{Should the discussion be merged with the result section?}

@@

In this contribution, we presented HealPixNet, a spherical CNN for the HealPix sampling. The gist of our construction resides in using a graph to model the sphere. 
This allows us to leverage the advantages of the graph convolution, i.e it is both efficient (with a complexity of $\mathcal{O}(n)$) and flexible (allow to work only on a subpart of the sphere) while it still captures well the spherical properties of the domain. 
We also show on a range of classification problems that the neural network systematically significantly outperform classical methods.

We provide HealPixNet as a small easy-to-use package available at \todo{\url{https://github.com/xxx}}. 
As a future work, we intend to 1) further investigate the link between the graph Fourier basis and the spherical harmonic and 2) compare HealPixNet to other spherical CNN using different sampling schemes.

@@

\section{Spherical harmonics VS graph Fourier eigenvectors}
\ref{fig:graph_harmonics} displays the first 16 eigenvectors of the
graph Fourier basis used for the convolution. While there is a clear
resemblance with the spherical harmonic, the small irregularities in the
sampling (see \figref{healpix_graph_4}) have an effect on the
graph Fourier eigenvectors. First, conterintuitively, some graph eigenvectors will be
localized~\cite{perraudin2018global}, i.e: they will span only a small subpart
of the sphere.  Second, as Nside increase toward infinity and hence the number
of pixels $n\rightarrow \infty$, we are still unsure if the graph eigenvectors
converge toward the spherical harmonics. We believe that it might be the case
for a different graph construction (fully connected) and that a proof could be
built on top of the work of~\cite{belkin2007convergence}. This is obviously
out of the scope of this contribution.

\label{sec:spherical_harmonics}
\begin{figure*}[!htb]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{The first 16 Fourier modes (eigenvectors of the graph Laplacian) of a graph constructed from the Healpix sampling of the sphere. Eigenvectors 1--3 could be associated with spherical harmonics of degree $\ell=1$, eigenvectors 4--8 with degree $\ell=2$, and eigenvectors 9--15 with degree $\ell=3$. Nevertheless, the graph eigenvectors are only approximating the real spherical harmonics.}
\label{fig:graph_harmonics}
\end{figure*}

@@

In \figref{gaussian_filters_visualization}, we present three different
visualization for a HealPix graph convolution kernel. The kernel used is defined in the graph spectral domain as $K_t(x)=e^{-\tau t x}$. This is the domain used for our first visualization. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. The lower eigenvalues $\lambda$ correspond to the lower frequencies, i.e. the spectral mode with less variation. 
As a second visualization, we can observe the kernel directly on the sphere. To do so, we perform the convolution of a
Kroneker on the sphere plot the result with gnomonic projection. Eventually, as we are
working on the sphere with theoretically spherical filters (up to the irregularities of the sampling), we could only plot the section of the convoluted Kroneker. To do so, we convolve a Kroneker on the equator and use the equator for our section as shown in~\figref{index_section}. Note that, because of the small irregularities in the HealPix
sampling, the second and the third methods are subject to small variations depending on the
chosen Kroneker.

@@

When comparing the network, we want to ensure that the SVM classifier has
access to the same amount of data that the network, i.e, potentially an
infinite number of samples. Hence, we fit different linear SVM classifier using
various training set size until we observe experimentally that
increasing the amount of data does not improve the classifier. The validation
set remain the same and is used for cross-validation to tune the L2
penalization parameter of SVM. Eventually, we evaluate the final classifier on
the testing set. An example of the evolution of the training and valdication
error is shown in \figref{hist_error_evolution}.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/hist_error_order2_noise1_5.pdf}
\caption{Evolution curve of the error with respect to the number of training samples for histogram features and a linear classifier (setting: order 2, relative noise level: 1.5). The validation error might be slightly below the training error because we use it for cross-validation.}
\label{fig:hist_error_evolution}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/border_effects.pdf}
\caption{Graph convolution of a Kroneker placed at two different spots on the graph. The graph build from $1/12$ of the sphere and an Nside of 16. When the Kroneker is placed on the left corner, the resulting convolution is not circular anymore due to border effect. The graph setting assumes reflective border conditions.}
\label{fig:border_effects}
\end{figure}
